Competing Bandits: The Perils of Exploration under Competition

### 2018-09-05 paper/content/inverted_u.tex
A partial pass.

We describe our main experiments with the duopoly game. On a high level, we demonstrate the inverted-U relationship between the algorithms incentivized by the competition algorithms and the intensity of the competition, from 
permanent monopoly to temporary monopoly to permanent duopoly.

@@

\begin{finding}
\textit{For sufficiently low warm start under the instances that are exploration-disadvantaged, the unique incentivized strategies in the competition game are:\footnote{The uniqueness comes from the fact that we break indifferences towards easier to deploy algorithms}
\begin{center}
\textbf{Permanent Monopoly} - $\DG$ \\
\textbf{Temporary Monopoly} (incumbent)- $\TS$ \\
\textbf{Permanent Duopoly} - $(\DG, \DG)$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{finding}

\xhdr{Permanent Monopoly.} Since there is only a single firm in the
market for the entire period, the firm can take the entire market
regardless of what algorithm it deploys. \swedit{Since we assume that exploration algorithms ($\DEG$ and $\TS$) would incur deployment cost and firms break indifferences towards easier to deploy algorithms}, the firm would choose to deploy $\DG$.

Next, to describe our results for temporary monopoly, we will
introduce the notion of \textit{effective end of game (EEOG)} --- the
last round $t$ such that the agent at round $(t-1)$ and the next agent
around at $t$ choose different firms.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\begin{table*}[h]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c|}{Heavy-Tail} 
  & \multicolumn{3}{c|}{Needle-in-Haystack}\\
  \hline 
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG 
  & \makecell{\textbf{0.29} $\pm$0.03\\ \Eeog 55 (0)} 
  & \makecell{\textbf{0.72} $\pm$0.02\\ \Eeog 570 (0)} 
  & \makecell{\textbf{0.76} $\pm$0.02\\ \Eeog 620 (99)} 
  %%
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 200 (27)}
  & \makecell{\textbf{0.6} $\pm$0.03\\ \Eeog 370 (0}
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 580 (122)}
  \\
\hline
  $\TS$ vs $\DEG$ 
  & \makecell{\textbf{0.3} $\pm$0.03\\ \Eeog 37 (0)} 
  & \makecell{\textbf{0.88} $\pm$0.01\\ \Eeog 480 (0)}
  & \makecell{\textbf{0.9} $\pm$0.01\\ \Eeog 570 (114)} 
  %%
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 150 (14}
  & \makecell{\textbf{0.52} $\pm$0.03\\ \Eeog 460 (79)}
  & \makecell{\textbf{0.56} $\pm$0.02\\ \Eeog 740 (628)} 
  \\
\hline
  $\DG$ vs $\DEG$ 
  & \makecell{\textbf{0.62} $\pm$0.03\\ \Eeog 410 (7)}
  & \makecell{\textbf{0.6} $\pm$0.02\\ \Eeog 790 (762)}
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 730 (608)} 
  %%
  & \makecell{\textbf{0.46} $\pm$0.03\\ \Eeog 340 (129)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 650 (408)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 690 (467)} 
  \\
   \hline
\end{tabular}
\caption{The duopoly experiment for Uniform and Needle-in-Haystack instances. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. The first line in the cell is the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. 
The second line specifies the ``effective end of game" (\Eeog): the average and the median (in brackets).}
\label{sim_table}
\end{table*}
\xhdr{Permanent Duopoly} \swedit{Since both firms enter at the same time, they will both get an expected market share of $1/2$ whenever they deploy
  the same algorithms. We will mainly investigate the market shares
  when the firms play different algorithms.  In Table \ref{sim_table}, we report the average market share taken over the
  $N$ simulations for different values of warm-up length $T_0$, and
  also the mean and median of the EEOG, under heavy tail and needle in
  haystack instances.}

For the exploration-disadvantaged instances, we see that, for low warm start, $\DG$ is a weakly dominant strategy for the competition game, and in particular weakly dominates $\TS$. This establishes our claim that $\DG$ is the incentivized algorithm.\footnote{We defer the table for uniform instances to the appendix. Summarizing the results, we see that, for low warm start, \DG yields more than half the market against \TS but is indifferent between \DG and \DEG. Since we break indifference towards easier to deploy algorithms, we also find that \DG is the incentivized algorithm in equilibrium}

@@

@@

\xhdr{Temporary Monopoly.} Recall that the incumbent firm enters the
market and serves as a monopolist until the entrant firm enters at
round $X$. \gaedit{For this section we focus on the warm start values and on the exploration-disadvantaged instances where we observed that, under permanent duopoly, $\DG$ was the weakly dominant strategy. Concretely, our presented results are for $T_0 = 20$.} \footnote{However, we report the results for the Needle In Haystack instances in the supplementary material. The results are unsurprising, where $\TS$ is the dominant
  strategy for both the entrant and incumbent.}

@@

\begin{table}[ht]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
   & $\TS$  & $\DEG$  & $\DG$ \\ \hline
$\TS$ 
    & \makecell{\textbf{0.003}$\pm$0.003} 
    & \makecell{\textbf{0.083}$\pm$0.02} 
    & \makecell{\textbf{0.17}$\pm$0.02} \\\hline
$\DEG$ 
    & \makecell{\textbf{0.045}$\pm$0.01} 
    & \makecell{\textbf{0.25}$\pm$0.02} 
    & \makecell{\textbf{0.23}$\pm$0.02} \\\hline
$\DG$ 
    & \makecell{\textbf{0.12}$\pm$0.02} 
    & \makecell{\textbf{0.36}$\pm$0.03} 
    & \makecell{\textbf{0.3}$\pm$0.02} \\\hline
\end{tabular}
\label{ht_incum}
\caption{The temporary monopoly experiment with $X=200$ (and $T_0=20$), for the Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the row) and the incumbent's algorithm (the column). The cell specifies the entrant's market share for the rounds in which he was present. We give the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\end{table}

>>>>

Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from permanent monopoly to temporary monopoly to permanent duopoly, we find a stylized inverted-U relationship as in Figure~\ref{fig:inverted-U}.

More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DG, \DEG and \TS. We do this for each "intensity level" and each MAB instance, and look for findings that are consistent across MAB instances.

For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs \cite{MWT-WhitePaper-2016,DS-arxiv}). Then \DG is trivially the dominant strategy under permanent monopoly.

%(1) Permanent monopoly - indifferent between all algorithms but break indifference towards DG cause of deployment costs
%(2) Temporary monopoly - TS is the dominant strategy for the incumbent
%(3) Permanent duopoly -
%   (a) Under NIH, TS is dominant
%   (b) Under Heavy Tail, DG is weakly dominant
%   (c) Under Uniform, DG "beats" TS, but DEG "beats" TS by more than DG and DEG vs DG leads to ~50% market share (it's slightly in favor of DG but 50/50 is within the confidence band). If we break indifference towards DG cause of deployment costs, then we have that the unique PSNE is (DG, DG). If we don't add the indifference breaking, then we have four PSNE, (DG, DEG), (DEG, DG), (DG, DG), (DEG, DEG)

For temporary monopoly, we find that \TS dominates:

\begin{finding}\label{find:temp-monopoly}
\textit{Under temporary monopoly, TS is the dominant strategy for the incumbent, across all MAB instances, if the temporary monopoly is sufficiently long-lasting.
}
\end{finding}

For permanent duopoly, it is a crucial whether an MAB instance is exploration-disadvantaged:

\begin{finding}\label{find:duopoly}
\textit{Under permanent duopoly: 
\begin{itemize}
\item[(a)] (\DG,\DG) is the unique pure-strategy Nash equilibrium for exploration-disadvantaged MAB instances with a sufficiently small ``warm start".
\item[(b)] This is not necessarily the case for MAB instances that are not exploration-disadvantaged. In particular, \TS is a weakly dominating strategy for Needle-in-Haystack.
\end{itemize}
}
\end{finding}

%\begin{finding}
%\textit{For sufficiently low warm start under the instances that are exploration-disadvantaged, the unique incentivized strategies in the competition game are:\footnote{The uniqueness comes from the fact that we break indifferences towards easier to deploy algorithms}
%\begin{center}
%\textbf{Permanent Monopoly} - $\DG$ \\
%\textbf{Temporary Monopoly} (incumbent)- $\TS$ \\
%\textbf{Permanent Duopoly} - $(\DG, \DG)$ \\
%\end{center}
%The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.}
%\end{finding}

%\xhdr{Permanent Monopoly.} Since there is only a single firm in the
%market for the entire period, the firm can take the entire market
%regardless of what algorithm it deploys. \swedit{Since we assume that exploration algorithms ($\DEG$ and $\TS$) would incur deployment cost and firms break indifferences towards easier to deploy algorithms}, the firm would choose to deploy $\DG$.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\begin{table*}[h]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail}
  & \multicolumn{3}{c|}{Needle-in-Haystack}\\
  \hline
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG
  & \makecell{\textbf{0.29} $\pm$0.03\\ \Eeog 55 (0)}
  & \makecell{\textbf{0.72} $\pm$0.02\\ \Eeog 570 (0)}
  & \makecell{\textbf{0.76} $\pm$0.02\\ \Eeog 620 (99)}
  %%
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 200 (27)}
  & \makecell{\textbf{0.6} $\pm$0.03\\ \Eeog 370 (0}
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 580 (122)}
  \\
\hline
  $\TS$ vs $\DEG$
  & \makecell{\textbf{0.3} $\pm$0.03\\ \Eeog 37 (0)}
  & \makecell{\textbf{0.88} $\pm$0.01\\ \Eeog 480 (0)}
  & \makecell{\textbf{0.9} $\pm$0.01\\ \Eeog 570 (114)}
  %%
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 150 (14}
  & \makecell{\textbf{0.52} $\pm$0.03\\ \Eeog 460 (79)}
  & \makecell{\textbf{0.56} $\pm$0.02\\ \Eeog 740 (628)}
  \\
\hline
  $\DG$ vs $\DEG$
  & \makecell{\textbf{0.62} $\pm$0.03\\ \Eeog 410 (7)}
  & \makecell{\textbf{0.6} $\pm$0.02\\ \Eeog 790 (762)}
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 730 (608)}
  %%
  & \makecell{\textbf{0.46} $\pm$0.03\\ \Eeog 340 (129)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 650 (408)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 690 (467)}
  \\
   \hline
\end{tabular}
\caption{Permanent duopoly, for Heavy-Tail and Needle-in-Haystack instances. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. Line 1 in the cell is the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
Line 2 specifies the ``effective end of game" (\Eeog): the average and the median (in brackets). The time horizon is $T=2000$.}
\label{sim_table}
\end{table*}

\subsubsection{Permanent Duopoly (Finding~\ref{find:duopoly}).} 
We investigate the firms' market shares when they choose different algorithms (otherwise, by symmetry both firms get half of the agents). We report the market shares for Heavy-Tail and Needle-in-Haystack instances in Table \ref{sim_table}  (see the first line in each cell), for a range of values of the warm start $T_0$.

We find that $\DG$ is a weakly dominant strategy for the Heavy-tail instance (with a sufficiently small $T_0$). However, \TS is a weakly dominating strategy for the Needle-in-Haystack instance.

A similar table for the Uniform MAB instance is deferred to the appendix. We find that for a sufficiently small $T_0$, \DG yields more than half the market against \TS,  but achieves similar market share vs. \DG and \DEG. By our tie-breaking rule, (\DG,\DG) is the only pure-strategy equilibrium.

@@

\OMIT{This establishes our claim that $\DG$ is the incentivized algorithm.\footnote{We defer the table for uniform instances to the appendix. Summarizing the results, we see that, for low warm start, \DG yields more than half the market against \TS but is indifferent between \DG and \DEG. Since we break indifference towards easier to deploy algorithms, we also find that \DG is the incentivized algorithm in equilibrium}}

@@

We introduce the notion of \textit{effective end of game (\Eeog)} the
last round $t$ such that the agents at this and previous round choose different firms.

@@

\subsubsection{Temporary Monopoly (Finding~\ref{find:temp-monopoly}).} 
Recall that the incumbent firm enters the
market and serves as a monopolist until the entrant firm enters at
round $X$. \gaedit{For this section we focus on the warm start values and on the exploration-disadvantaged instances where we observed that, under permanent duopoly, $\DG$ was the weakly dominant strategy. Concretely, our presented results are for $T_0 = 20$.} \footnote{However, we report the results for the Needle In Haystack instances in the supplementary material. The results are unsurprising, where $\TS$ is the dominant
  strategy for both the entrant and incumbent.}

@@

\begin{table}[ht]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
   & $\TS$  & $\DEG$  & $\DG$ \\ \hline
$\TS$
    & \makecell{\textbf{0.003}$\pm$0.003}
    & \makecell{\textbf{0.083}$\pm$0.02}
    & \makecell{\textbf{0.17}$\pm$0.02} \\\hline
$\DEG$
    & \makecell{\textbf{0.045}$\pm$0.01}
    & \makecell{\textbf{0.25}$\pm$0.02}
    & \makecell{\textbf{0.23}$\pm$0.02} \\\hline
$\DG$
    & \makecell{\textbf{0.12}$\pm$0.02}
    & \makecell{\textbf{0.36}$\pm$0.03}
    & \makecell{\textbf{0.3}$\pm$0.02} \\\hline
\end{tabular}
\label{ht_incum}
\caption{Temporary monopoly, with $X=200$ (and $T_0=20$), for the Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the row) and the incumbent's algorithm (the column). The cell specifies the entrant's market share (fraction of rounds in which it was chosen) for the rounds in which he was present. We give the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\end{table}
